BOLD Parameter Estimation Using Particle Filters

### 2010-12-13 doc/thesis/background.tex
updated thesis for first ETD

\begin{figure}
\centering
\includegraphics[scale=.7]{images/HRF}
\caption[Canonical Hemodynamic Response Function]
{Canonical Hemodynamic Response Function, normalization makes $y$ units
arbitrary}
\label{fig:HRF}
\end{figure}

@@

\begin{figure}
\includegraphics[width=16cm]{images/kalman2}
\caption[Example Kalman Filter Progression]
{a) Sigma Points are passed through the state/measurement equations
to get a state prediction (green line). The first two moments of
the state prediction are then used to estimate the distribution with
a Gaussian (red line). The distribution centered at the new measurement,
estimating measurement error (dark blue), is then combined with the existing
state prediction (red line) to get the new estimated state (light blue).
The same process is applied in b), and c) with different measurements and
initial state estimates.}
\label{fig:EKFWorking}
\end{figure}

The difficulty
of using a Kalman Filter, however, is that it assumes a multivariate 
Gaussian for the state variables, $X(t-1)$. The more nonlinear the system
gets the more likely that the Gaussian will be insufficient to describe
the distribution, $X(t)$. When this occurs, every step from  $X(t-1)$ 
to $X(t)$ will introduce additional
error in the posterior distribution. Furthermore, it is not really known what 
sort of underlying distributions may exist in such a mixed biological,
mechanical, chemical system such as the brain.
 The assumption of Gaussianity is what allows
the \ac{UKF} to estimate the posterior distribution using only the 
first and second moments;
however, if this assumption is violated significant error will result.
 On the other hand, for
small variances and short time steps the Gaussian distribution is a good 
fit, and so in some cases the Unscented Kalman Filter could work quite
well. 
\begin{table}[t]
\centering
\begin{tabular}{|c || c |}
\hline 
Parameter & Run 1 \\
\hline
$\tau_0$ & .98  \\
$\alpha$ & .33 \\
$E_0$ & .34  \\
$V_0$ & .03  \\
$\tau_s$ & 1.54  \\
$\tau_f$ & 2.46  \\
$\epsilon$ & .54  \\
$V_t$ & N(1, .09)  \\
$Q_t$ & N(1, .09)  \\
$S_t$ & N(1, .09) \\
$F_t$ & N(1, .09) \\
\hline
\end{tabular}
\caption{Parameters used to test Gaussianity of variables after being transitioned through
the \ac{BOLD} model}
\label{tab:steptable} 
\end{table}

>>>>

\begin{figure}
\centering
\includegraphics[scale=.7]{images/HRF}
\caption[Canonical Hemodynamic Response Function]
{Canonical Hemodynamic Response Function. Normalization makes $y$ units
arbitrary}
\label{fig:HRF}
\end{figure}

@@

\begin{figure}
\includegraphics[width=16cm]{images/kalman2}
\caption[Example Kalman Filter Progression]
{Example Kalman Filter Progression. a) Sigma Points are passed through the state/measurement equations
to get a state prediction (green line). The first two moments of
the state prediction are then used to estimate the distribution with
a Gaussian (red line). The distribution centered at the new measurement,
estimating measurement error (dark blue), is then combined with the existing
state prediction (red line) to get the new estimated state (light blue).
The same process is applied in b), and c) with different measurements and
initial state estimates.}
\label{fig:EKFWorking}
\end{figure}

The difficulty
of using a Kalman Filter, however, is that it assumes a multivariate 
Gaussian for the state variables, $X(t-1)$. The more nonlinear the system
gets the more likely that the Gaussian will be insufficient to describe
the distribution, $X(t)$. When this occurs, every step from  $X(t-1)$ 
to $X(t)$ will introduce additional
error in the posterior distribution. Furthermore, it is not really known what 
sort of underlying distributions may exist in such a mixed biological,
mechanical, chemical system such as the brain.
 The assumption of Gaussianity is what allows
the \ac{UKF} to estimate the posterior distribution using only the 
first and second moments;
however, if this assumption is violated significant error will result.
 On the other hand, for
small variances and short time steps the Gaussian distribution is a good 
fit, and so in some cases the Unscented Kalman Filter could work quite
well. 
\begin{table}[t]
\centering
\begin{tabular}{|c || c |}
\hline 
Parameter & Run 1 \\
\hline
$\tau_0$ & .98  \\
$\alpha$ & .33 \\
$E_0$ & .34  \\
$V_0$ & .03  \\
$\tau_s$ & 1.54  \\
$\tau_f$ & 2.46  \\
$\epsilon$ & .54  \\
$V_t$ & N(1, .09)  \\
$Q_t$ & N(1, .09)  \\
$S_t$ & N(1, .09) \\
$F_t$ & N(1, .09) \\
\hline
\end{tabular}
\caption{Parameters used to test Gaussianity of variables after being transitioned through
the \ac{BOLD} model.}
\label{tab:steptable} 
\end{table}

### 2010-12-13 doc/thesis/intro.tex
updated thesis for first ETD

\begin{figure}
\centering
\includegraphics[width=.5\textwidth]{images/bold}
\caption[Diagram of BOLD Signal]{A) End of Pre-Stimulus Undershoot, $< 2s$ typical,
B) End of Stimulus, C) Input Bloodflow Drops from Supernormal to Subnormal,
$5 s$ after B typical, D) Beginning of Post-stimulus Undershoot, $12 s$ after B
typical, E) Full return to normal, between $20-30s$ after B. \cite{Behzadi2005}}
\end{figure}

\section{\ac{fMRI}}
\ac{MRI}, is a method of building 3D images
non-invasively, based on the difference between nuclear spin
relaxation times in different molecules. The technique is
well established and documented in various works \cite{Bushberg2002}. 
First, the subject 
is brought into a large magnetic field which causes nuclear spins
to align. \ac{RF} signals may
then be used to excite nuclear spin away from the base alignment. 
As the nuclei precess back to the alignment of the magnetic
field, they emit detectable \ac{RF} signals. Conveniently, the
nuclear spins return their original state at different
rates, called the \ac{T1} relaxation time, depending on the material excited.
Additionally, the
coherence of the spins also decays differently (and roughly an order of 
magnitude faster
than \ac{T1} relaxation) based on the properties of the region 
\cite{Bushberg2002}.
This gives two primary methods of contrasting substances,
which form the basis of \ac{T1} and \ac{T2} weighted images. Additionally, 
dephasing occurs at two different rates, the \ac{T2} relaxation time,
which is unrecoverable, and \ac{T2}$^*$ relaxation, which is
much faster, but possible to recover from via special \ac{RF} signals.
\ac{T1} relaxation times are typically on the order of seconds if 
a sufficiently strong excitation is applied, whereas \ac{T2}$^*$ relaxation
times are usually less than 100ms. 
In order to rapidly acquire entire brain images, as done in Functional 
\ac{MRI}, a single large excitation pulse is applied to the entire brain,
and the entire volume is acquired in a single \ac{T1} relaxation period
\cite{Obata2004}. 
Because the entire k-space (spatial-frequency) volume is acquired 
from a single excitation, the \ac{SNR} is very low
in \ac{EPI} \cite{Smith2007}.

@@

\section{Properties of the \ac{BOLD} Model}
\label{sec:BOLD Analysis}
Since the first complete \ac{BOLD} model was proposed by Friston et al. 
\cite{Friston2000},  
several studies have analyzed its properties.  
The most important property is that the system is dissipative, and given
enough time will converge to a constant value. This is found simply by
analyzing the eigenvalues of the state equation Jacobian, 
\cite{Deneux2006, Hu2009}. The steady state of the Balloon
model equations gives:

@@

%note to self, friston2002b's parameters are from a picture
\begin{table}[t]
\centering
\begin{tabular}{|c || c | c | c | c|}
\hline 
Parameter  & Friston et al. \cite{Friston2000} & Jonston et al. \cite{Johnston2007} 
    & Vakorin et al. \cite{Vakorin2007} & Deneux et al. \cite{Deneux2006}\\
\hline
$\tau_0  $ &  $N(.98 , .25^2)$  & $8.38 \pm 1.5  $ & $.94$ & .27\\
$\alpha  $ &  $N(.33 , .045^2)$  & $.189 \pm .004 $ & $.4$ (NC) & .63 \\
$E_0     $ &  $N(.34 , .1 ^2)$  & $.635 \pm .072 $ & $.6$ (NC) & .33\\
$V_0     $ &  $.03$ (NC)        & $.0149 \pm .006$ & (NC) & .16\\
$\tau_s  $ &  $N(1.54, .25^2)$  & $4.98 \pm 1.07 $ & $2.2$ & 2.04 \\
$\tau_f  $ &  $N(2.46, .25^2)$  & $8.31 \pm 1.51 $ & $.45$ & 5.26\\
$\epsilon$ &  $N(.54 , .1 ^2)$  & $.069 \pm .014 $ & (NC) & .89\\
\hline
\end{tabular}
\caption{Parameters found in various studies. (NC) indicates that the value
wasn't calculated. \cite{Vakorin2007} made use of the values from \cite{Friston2000}
where not explicitly stated}
\label{tab:Params} 
\end{table}

>>>>

\begin{figure}
\centering
\includegraphics[width=.5\textwidth]{images/bold2}
\caption[Diagram of \ac{BOLD} Signal]{Diagram of the \ac{BOLD} Sigmal.
A) End of Pre-Stimulus Undershoot, $< 2s$ typical,
B) End of Stimulus, C) Input Bloodflow Drops from Supernormal to Subnormal,
$5 s$ after B typical, D) Beginning of Post-stimulus Undershoot, $12 s$ after B
typical, E) Full return to normal, between $20-30s$ after B.}
\end{figure}

\section{Functional Magnetic Resonance Imaging}
\ac{MRI}, is a method of building 3D images
non-invasively, based on the difference between nuclear spin
relaxation times in different molecules. The technique is
well established and documented in various works \cite{Bushberg2002}. 
First, the subject 
is brought into a large magnetic field which causes nuclear spins
to align. \ac{RF} signals may
then be used to excite nuclear spin away from the base alignment. 
As the nuclei precess back to the alignment of the magnetic
field, they emit detectable \ac{RF} signals. Conveniently, the
nuclear spins return their original state at different
rates, called the \ac{T1} relaxation time, depending on the material excited.
Additionally, the
coherence of the spins also decays differently (and roughly an order of 
magnitude faster
than \ac{T1} relaxation) based on the properties of the region 
\cite{Bushberg2002}.
This gives two primary methods of contrasting substances,
which form the basis of \ac{T1} and \ac{T2} weighted images. Additionally, 
dephasing occurs at two different rates, the \ac{T2} relaxation time,
which is unrecoverable, and \ac{T2}$^*$ relaxation, which is
much faster, but possible to recover from via special \ac{RF} signals.
\ac{T1} relaxation times are typically on the order of seconds if 
a sufficiently strong excitation is applied, whereas \ac{T2}$^*$ relaxation
times are usually less than 100ms. 
In order to rapidly acquire entire brain images, as done in Functional 
\ac{MRI}, a single large excitation pulse is applied to the entire brain,
and the entire volume is acquired in a single \ac{T1} relaxation period
\cite{Obata2004}. 
Because the entire k-space (spatial-frequency) volume is acquired 
from a single excitation, the \ac{SNR} is very low
in \ac{EPI} \cite{Smith2007}.

@@

\section{Properties of the Blood-Oxygen-Level Dependent Model}
\label{sec:BOLD Analysis}
Since the first complete \ac{BOLD} model was proposed by Friston et al. 
\cite{Friston2000},  
several studies have analyzed its properties.  
The most important property is that the system is dissipative, and given
enough time will converge to a constant value. This is found simply by
analyzing the eigenvalues of the state equation Jacobian, 
\cite{Deneux2006, Hu2009}. The steady state of the Balloon
model equations gives:

@@

%note to self, friston2002b's parameters are from a picture
\begin{table}[t]
\centering
\begin{tabular}{|c || c | c | c | c|}
\hline 
Parameter  & Friston et al. \cite{Friston2000} & Jonston et al. \cite{Johnston2007} 
    & Vakorin et al. \cite{Vakorin2007} & Deneux et al. \cite{Deneux2006}\\
\hline
$\tau_0  $ &  $N(.98 , .25^2)$  & $8.38 \pm 1.5  $ & $.94$ & .27\\
$\alpha  $ &  $N(.33 , .045^2)$  & $.189 \pm .004 $ & $.4$ (NC) & .63 \\
$E_0     $ &  $N(.34 , .1 ^2)$  & $.635 \pm .072 $ & $.6$ (NC) & .33\\
$V_0     $ &  $.03$ (NC)        & $.0149 \pm .006$ & (NC) & .16\\
$\tau_s  $ &  $N(1.54, .25^2)$  & $4.98 \pm 1.07 $ & $2.2$ & 2.04 \\
$\tau_f  $ &  $N(2.46, .25^2)$  & $8.31 \pm 1.51 $ & $.45$ & 5.26\\
$\epsilon$ &  $N(.54 , .1 ^2)$  & $.069 \pm .014 $ & (NC) & .89\\
\hline
\end{tabular}
\caption[Parameters found in various studies.]{Parameters found in various studies. (NC) indicates that the value
wasn't calculated. \cite{Vakorin2007} made use of the values from \cite{Friston2000}
where not explicitly stated}
\label{tab:Params} 
\end{table}

### 2010-12-13 doc/thesis/methods.tex
more on isbi
updated thesis for first ETD

\begin{figure}
\centering
\subfigure[]{\label{fig:QQDDelta:A}\includegraphics[trim=6cm 1cm 6cm 1cm,width=13cm]{images/noise2_0009d_29_49_9}}
\subfigure[]{\label{fig:QQDDelta:B}\includegraphics[trim=6cm 1cm 6cm 1cm,width=13cm]{images/noise2_0009d_34_43_24}}
\subfigure[]{\label{fig:QQDDelta:C}\includegraphics[trim=6cm 1cm 6cm 1cm,width=13cm]{images/noise2_0009d_22_38_23}}
\subfigure[]{\label{fig:QQDDelta:D}\includegraphics[trim=6cm 1cm 6cm 1cm,width=13cm]{images/noise2_0009d_37_29_24}}
\caption[Resting State Noise Analysis, Steps]{Resting state noise analysis, where
signal levels are the difference between adjacent signals in \autoref{fig:QQDC}.
From Left to Right:
(resting state) signal, Q-Q Plot of measurements with the standard Normal,
Signal auto-regression, Histogram of the Points}
\label{fig:QQDelta}
\end{figure}

\begin{figure}
\centering
\subfigure[]{\label{fig:QQs:A}\includegraphics[trim=6cm 1cm 6cm 1cm,width=13cm]{images/noise2_0009s_29_49_9}}
\subfigure[]{\label{fig:QQs:B}\includegraphics[trim=6cm 1cm 6cm 1cm,width=13cm]{images/noise2_0009s_34_43_24}}
\subfigure[]{\label{fig:QQs:C}\includegraphics[trim=6cm 1cm 6cm 1cm,width=13cm]{images/noise2_0009s_22_38_23}}
\subfigure[]{\label{fig:QQs:D}\includegraphics[trim=6cm 1cm 6cm 1cm,width=13cm]{images/noise2_0009s_37_29_24}}
\caption[Resting State Noise Analysis, After Subtracting Spline]
{Resting State Noise Analysis, where
signal levels are the result of subtracting a spline from the signal in 
\autoref{fig:QQDC}.
From Left to Right:
(resting state) signal, Q-Q Plot of measurements with the standard Normal,
Signal auto-regression, Histogram of the Points}
\label{fig:QQSpline}
\end{figure}

>>>>

\begin{figure}
\centering
\subfigure[]{\label{fig:QQDDelta:A}\includegraphics[trim=6cm 1cm 6cm 1cm,width=13cm]{images/noise2_0009d_29_49_9}}
\subfigure[]{\label{fig:QQDDelta:B}\includegraphics[trim=6cm 1cm 6cm 1cm,width=13cm]{images/noise2_0009d_34_43_24}}
\subfigure[]{\label{fig:QQDDelta:C}\includegraphics[trim=6cm 1cm 6cm 1cm,width=13cm]{images/noise2_0009d_22_38_23}}
\subfigure[]{\label{fig:QQDDelta:D}\includegraphics[trim=6cm 1cm 6cm 1cm,width=13cm]{images/noise2_0009d_37_29_24}}
\caption[Resting state noise analysis, where signal levels are the difference between adjacent signals in \autoref{fig:QQDC}]{Resting state noise analysis, where
signal levels are the difference between adjacent signals in \autoref{fig:QQDC}.
From Left to Right:
(resting state) signal, Q-Q Plot of measurements with the standard Normal,
Signal auto-regression, Histogram of the Points.}
\label{fig:QQDelta}
\end{figure}

\begin{figure}
\centering
\subfigure[]{\label{fig:QQs:A}\includegraphics[trim=6cm 1cm 6cm 1cm,width=13cm]{images/noise2_0009s_29_49_9}}
\subfigure[]{\label{fig:QQs:B}\includegraphics[trim=6cm 1cm 6cm 1cm,width=13cm]{images/noise2_0009s_34_43_24}}
\subfigure[]{\label{fig:QQs:C}\includegraphics[trim=6cm 1cm 6cm 1cm,width=13cm]{images/noise2_0009s_22_38_23}}
\subfigure[]{\label{fig:QQs:D}\includegraphics[trim=6cm 1cm 6cm 1cm,width=13cm]{images/noise2_0009s_37_29_24}}
\caption[Resting State Noise Analysis, where
signal levels are the result of subtracting a spline from the signal in 
\autoref{fig:QQDC}]
{Resting State Noise Analysis, where
signal levels are the result of subtracting a spline from the signal in 
\autoref{fig:QQDC}.
From Left to Right:
(resting state) signal, Q-Q Plot of measurements with the standard Normal,
Signal auto-regression, Histogram of the Points}
\label{fig:QQSpline}
\end{figure}

### 2010-12-13 doc/thesis/particlefilters.tex
added some more images. and about half done with ISBI
updated thesis for first ETD

Additionally, for the particle filter $y_k$ and $y_{0:k-1}$ are 
constant across all particles, thus $P(y_k| y_{0:k-1})$ can
be dropped when the equality is changed to a proportion. 
Using these properties, $P(x^i_{0:k} | y_{0:k})$ may be broken up as follows 
(primarily using Bayes' Theorem):
\begin{equation}
\begin{array}{lclr}
P(x_{0:k} | y_{0:k}) & = & \frac{P(y_{0:k}, x_{0:k})}{P(y_{0:k})} & \\
 & = & \frac{P(y_k, x_{0:k} | y_{0:k-1}) \cancel{P(y_{0:k-1})}}{P(y_k | y_{0:k-1}) \cancel{P(y_{0:k-1})}} & \\
 & = & \frac{P(y_k| x_{0:k}, y_{0:k-1}) P(x_{0:k} | y_{0:k-1})}{P(y_k | y_{0:k-1}) } & \\
 & = & \frac{P(y_k| x_{0:k}, y_{0:k-1}) P(x_k | x_{0:k-1}, y_{0:k-1}) P(x_{0:k-1} | y_{0:k-1})}{P(y_k | y_{0:k-1})} &  \\
& = & \frac{P(y_k| x_k) P(x_k | x_{k-1}) P(x_{0:k-1} | y_{0:k-1})}{P(y_k | y_{0:k-1})}  & [\text{\autoref{eq:MarkovProperty}}]\\
& \propto & P(y_k| x_k) P(x_k | x_{k-1}) P(x_{0:k-1} | y_{0:k-1}) & [P(y_k|y_{0:k-1}) \text{ is constant}]
 \end{array}
 \label{eq:UpdateBayes}
\end{equation}
Inserting \autoref{eq:ImportanceDensity} and the result of \autoref{eq:UpdateBayes}
into \autoref{eq:weightfunc} leads to:
\begin{eqnarray}
w^i_k & \propto & \frac{P(y_k| x^i_k) \cancel{P(x^i_k | x^i_{k-1})} P(x^i_{0:k-1} | y_{0:k-1})}
                         {\cancel{P(x^i_k | x^i_{k-1})}q(x^i_{0:k-1} | y_{0:k-1})} \nonumber \\
& \propto & w^i_{k-1}P(y_k| x_k) 
\label{eq:weightevolve}
\end{eqnarray}

@@

\section{Sequential Importance Resampling}
\label{sec:Particle Filter Resampling}
\begin{figure}
\includegraphics[width=16cm]{images/particle_filter2}
\caption[Example Particle Filter Progression]{a) Initial particle locations
uniformly distributed in [-10, 10], with
weights set to give the State Prediction (green). When a measurement is
received it converted to a distribution approximating uncertainty. This
distribution is then used to alter the weights of the existing particles.
b) Another measurement is applied similar to a), c) Resampling draws
new particles in locations according to the Mixture \ac{PDF} approximated 
by the particle weights. d) Another measurement is applied as before,
but now with non-uniform particle locations.}
\end{figure}

>>>>

Additionally, for the particle filter $y_k$ and $y_{0:k-1}$ are 
constant across all particles, thus $P(y_k| y_{0:k-1})$ can
be dropped when the equality is changed to a proportion. 
Using these properties, $P(x^i_{0:k} | y_{0:k})$ may be broken up as follows 
(primarily using Bayes' Theorem):
\begin{equation}
\begin{array}{lclr}
P(x_{0:k} | y_{0:k}) & = & \frac{P(y_{0:k}, x_{0:k})}{P(y_{0:k})} & \\
 & = & \frac{P(y_k, x_{0:k} | y_{0:k-1}) \cancel{P(y_{0:k-1})}}{P(y_k | y_{0:k-1}) \cancel{P(y_{0:k-1})}} & \\
 & = & \frac{P(y_k| x_{0:k}, y_{0:k-1}) P(x_{0:k} | y_{0:k-1})}{P(y_k | y_{0:k-1}) } & \\
 & = & \frac{P(y_k| x_{0:k}, y_{0:k-1}) P(x_k | x_{0:k-1}, y_{0:k-1}) P(x_{0:k-1} | y_{0:k-1})}{P(y_k | y_{0:k-1})} &  \\
& = & \frac{P(y_k| x_k) P(x_k | x_{k-1}) P(x_{0:k-1} | y_{0:k-1})}{P(y_k | y_{0:k-1})}  & [\text{\autoref{eq:MarkovProperty}}]\\
& \propto & P(y_k| x_k) P(x_k | x_{k-1}) P(x_{0:k-1} | y_{0:k-1}) & [P(y_k|y_{0:k-1}) \text{ is constant}]
 \end{array}
 \label{eq:UpdateBayes}
\end{equation}
Inserting \autoref{eq:ImportanceDensity} and the result of \autoref{eq:UpdateBayes}
into \autoref{eq:weightfunc} leads to:
\begin{eqnarray}
w^i_k & \propto & \frac{P(y_k| x^i_k) \cancel{P(x^i_k | x^i_{k-1})} P(x^i_{0:k-1} | y_{0:k-1})}
                         {\cancel{P(x^i_k | x^i_{k-1})}q(x^i_{0:k-1} | y_{0:k-1})} \nonumber \\
& \propto & w^i_{k-1}P(y_k| x^i_k) 
\label{eq:weightevolve}
\end{eqnarray}

@@

\section{Sequential Importance Resampling}
\label{sec:Particle Filter Resampling}
\begin{figure}
\includegraphics[width=16cm]{images/particle_filter2}
\caption[Example Particle Filter Progression.]{
Example Particle Filter Progression. a) Initial particle locations
uniformly distributed in [-10, 10], with
weights set to give the State Prediction (green). When a measurement is
received it converted to a distribution approximating uncertainty. This
distribution is then used to alter the weights of the existing particles.
b) Another measurement is applied similar to a), c) Resampling draws
new particles in locations according to the Mixture \ac{PDF} approximated 
by the particle weights. d) Another measurement is applied as before,
but now with non-uniform particle locations.}
\end{figure}
